Predicting Success of Clinical Interviews via Probabilistic Modeling of Patient-Provider Communication Sequences

### 2017-03-09 paper/AMIA/main.tex
update table 5

\begin{table}[h]
\centering
\caption{Statistics of experimental dataset.}
\label{tab:data_dist}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Sequence type} & \textbf{Total sequences}  & \textbf{\# Successful sequences}  & \textbf{\# Unsuccessful sequences} & \textbf{Avg. length} \\ \hline      
 natural sequences & 1072 & 880 (82.09\%) & 192 (17.91\%) & 5.83 \\\hline
alternating sequences & 18753 & 18139 (96.72\%) & 614 (3.28\%) & 12.89 \\\hline 
  \end{tabular}
\end{table}

@@

\begin{equation}
p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N p_M(c_i|c_{i-1})-\sum_{i=2}^N p_{\overline M}(c_i|c_{i-1})\label{eq:class}
\end{equation}

@@

\textbf {Hidden Markov Model (HMM)} is a powerful tool for statistical modeling of processes varying in time. HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence. HMM can be considered as a doubly embedded stochastic process with a process that is not observable (hidden process) and can only be observed through another stochastic process (observable process) that produces a sequence of observations. An HMM can be fully specified by the following quintuple:

\begin{center}
$\lambda = (N, M, A, B, \pi)$
\end{center}

@@

We exclude the structure parameters M and N to designate HMM using a compact notation:

\begin{center}
$\lambda = (A, B, \pi)$
\end{center}

The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model figures out a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state. The Baum-Welch algorithm was used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm was used to determine the most likely sequence of hidden states for a given sequence of observations. After assignment of hidden states, the log-likelihood of successful outcome can be estimated using Eq.~\ref{eq:class}. As in the case of MC, we also experimented with second-order HMM.

@@

\subsection*{\textit{Predictive performance for alternating PPC code sequences}}
Table~\ref{tab:result_alt_seq} summarizes predictive performance of the proposed methods on alternating sequences. These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall. Similar to the case of natural sequences, first-order MC has the lowest performance compared to all other models across all metrics. Also, as in the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 and the highest accuracy of 0.9713 among all models. However, different from natural sequences, second-order MC demonstrates the highest precision 0.9778 as well as the highest F-measure of 0.9736. We also observed that second-order MC and HMM outperform first-order MC and HMM. Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively. Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 4.82\%, respectively.

@@

 \multirow{2}{*}{Markov Chain} & First order & 0.9777 & 0.9437 & 0.9604 & 0.9437\\\cline{2-6}
 & Second order & \textbf{0.9778} & 0.9694 & \textbf{0.9736} & 0.9694\\ \hline
 \multirow{2}{*}{Hidden Markov Model} & First order & 0.9392 & 0.9313 & 0.9253 & 0.9392\\ \cline{2-6}
 & Second order & 0.9704 & \textbf{0.9713} & 0.9699  & \textbf{0.9713}\\ \hline

\subsection*{\textit{Common patterns}}
Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews. It can be seen that most successful patterns start with a summary of the discussion or open-ended/close-ended question. After that, if adolescents show positive behavior change, the counselor immediately reflects on that positive tendency, to reinforce adolescent's desire to adhere to positive behavior change. On the other hand, general information by the counselor can lead to a negative behavior change in adolescents, even in the cases when adolescents were showing positive tendency in their previous communication. This observation can be explained by adolescents quickly loosing focus when presented with general information.

\begin{table}[h]
\centering
\caption{Common patterns in successful and unsuccessful motivational interviews.}
\label{tab:common_patterns}
  \begin{tabular}{|l|l|l|l|l|l|}
  \hline
   \textbf{Type} & \textbf{Pattern} \\ \hline      
successful & 328 SUM-S: Summarize $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 313 R-CML+: Reflect, \\ 
& commitment language positive $\rightarrow $ [positive commitment] \\\hline
successful & 307 SPT: Support $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 313 R-CML+: Reflect, \\
& commitment language positive $\rightarrow $ [positive commitment] \\\hline
successful & 306 CQ-EF: Closed question, Elicit Feedback $\rightarrow $ 120 HUP-O: High Uptake, other \\
& $\rightarrow $ 313 R-CML+: Reflect, commitment language positive $\rightarrow $ [positive commitment] \\\hline
unsuccessful & 311 R-CT+: Reflect, change talk positive $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 302 G-INFO+:  \\
& General Information, positive $\rightarrow $ [negative commitment] \\\hline
unsuccessful & 302 G-INFO+: General Information, positive $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $  \\
& 302 G-INFO+: General Information, positive $\rightarrow $ [negative commitment] \\\hline
unsuccessful & 305 EA: Emphasize Autonomy $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 302 G-INFO+:  \\
& General Information, positive $\rightarrow $ [negative commitment] \\\hline
  \end{tabular}
\end{table}

>>>>

\begin{table}[h]
\centering
\caption{Statistics of experimental dataset.}
\label{tab:data_dist}
  \begin{tabular}{|l|l|l|l|l|}
  \hline
   \textbf{Sequence type} & \textbf{Total sequences}  & \textbf{\# Successful sequences}  & \textbf{\# Unsuccessful sequences} & \textbf{Avg. length} \\ \hline      
 natural sequences & 1072 & 880 (82.09\%) & 192 (17.91\%) & 5.83 \\\hline
alternating sequences & 18753 & 18139 (96.72\%) & 614 (3.28\%) & 12.89 \\\hline 
  \end{tabular}
\end{table}

@@

\begin{equation}
p(S\rightarrow CML+) = \log\left(\frac{P_M(S)}{P_{\overline M}(S)}\right)= \sum_{i=2}^N \log p_M(c_i|c_{i-1})-\sum_{i=2}^N \log p_{\overline M}(c_i|c_{i-1})\label{eq:class}
\end{equation}

@@

\textbf {Hidden Markov Model (HMM)} is a powerful tool for statistical modeling of processes varying in time. HMMs are widely used for sequence analysis because of their ability to incorporate dependencies among elements in sequence. HMM can be considered as a doubly embedded stochastic process with a process that is not observable (hidden process) and can only be observed through another stochastic process (observable process) that produces a sequence of observations. An HMM can be fully specified by the following quintuple: $\lambda = (N, M, A, B, \pi)$

@@

We exclude the structure parameters M and N to designate HMM using a compact notation: $\lambda = (A, B, \pi)$. The key difference between HMM and MC is that HMM requires specifying the number of hidden states as a model parameter and then the model figures out a sequence of hidden states that best explains the observations along with state transition probabilities and distributions of observation symbols for each state. The Baum-Welch algorithm was used to estimate the parameters of HMMs for successful and unsuccessful interviews using the corresponding training set, while the Viterbi algorithm was used to determine the most likely sequence of hidden states for a given sequence of observations. After assignment of hidden states, the log-likelihood of successful outcome can be estimated using Eq.~\ref{eq:class}. As in the case of MC, we also experimented with second-order HMM.

@@

\subsection*{\textit{Predictive performance for alternating PPC code sequences}}
Table~\ref{tab:result_alt_seq} summarizes predictive performance of the proposed methods on alternating sequences. These results indicate that second-order HMM had better recall than MC, achieving 97.13\% recall. Similar to the case of natural sequences, first-order MC has the lowest performance compared to all other models across all metrics. Also, as in the case of natural sequences, prediction method based on second-order HMM has the highest recall of 0.9713 and the highest accuracy of 0.9713 among all models. However, different from natural sequences, second-order MC demonstrates the highest precision 0.9778 as well as the highest F-measure of 0.9736. We also observed that second-order MC and HMM outperform first-order MC and HMM. Specifically, moving from first-order to second-order MC improves precision, recall and F-measure of the MI success prediction method by 0.01\%, 2.72\% and 1.37\%, respectively. Moving from first-order to second-order HMM improves precision, recall and F-measure of the MI success prediction method by 3.32\%, 4.30\% and 3.70\%, respectively.\\

@@

 \multirow{2}{*}{Markov Chain} & First order & 0.9777 & 0.9437 & 0.9604 & 0.9437\\\cline{2-6}
 & Second order & \textbf{0.9778} & 0.9694 & \textbf{0.9736} & 0.9694\\ \hline
 \multirow{2}{*}{Hidden Markov Model} & First order & 0.9392 & 0.9313 & 0.9353 & 0.9392\\ \cline{2-6}
 & Second order & 0.9704 & \textbf{0.9713} & 0.9699  & \textbf{0.9713}\\ \hline
  \end{tabular}
\end{table}

\subsection*{\textit{Impact of hidden states}}
Since the number of hidden states may impact the classification
performance, we use an optimum number of hidden states in HMMs. Figure~\ref{fig:hidden-states} shows the performance of HMMs with a different number of hidden states. The results illustrate that, in general, the state number didn't impact the classification performance significantly. All four models show their best performance when state number in between 4-8. Table~\ref{tab:emission_matrix} shows an example of the distribution of PPC codes over the states for successful (S) and unsuccessful (U) HMMs. Log probability of each code reflects the state significance. In other words, these probabilities indicate that which PPC codes are close to adolescent commitment language and which are not?

\begin{figure}[htb!]
    \centering
    \includegraphics[width=0.70\textwidth]{figures/hidden-states.png}
    \caption{\textbf{Performance of HMMs with different sequence type, order, and number of hidden states used in the experiment}.}
    \label{fig:hidden-states}
\end{figure}

\begin{table}[h]
\centering
\caption{Distribution of PPC codes over the states for successful (S) and unsuccessful (U) HMMs. Significant PPC codes are given in short form, which have log probability higher than -2.}
\label{tab:emission_matrix}
  \begin{tabular}{|l|l|}
  \hline
   \textbf{State} & \textbf{Log probability of each symbol (301,LUP+, 327,OQ-TBN,101,EA,R-CT+,OQ-ECT+,AR,SUM-S,} \\ 
   \textbf{} & \textbf{R-CT-S,RBA-S,G-INFO+,AF,314,CQ-EF,119,R-CML+, OQ-ECML+, 307, 325, 344,CHT-,106,} \\
    & \textbf{OQ-ECT-, HUP-O, 345, 309, R-CT-S, 114, 304, 115, 329, 330, 103, 116, 324, 323, 118, 303, 334)} \\\hline      
S-1 & -2.896, -4.595, -4.031, -5.584, -11.383, -2.772, -1.532, -2.78, -3.816, -2.745, -8.444, -4.42, -3.485, \\
& -2.438, -8, -5.577, -10.333, -2.216, -3.462, -3.479, -3.71, -5.615, -20.984, -7.635, -5.7, -5.694, -3.629, \\
& -3.32, -3.195, -8, -5.007, -8, -4.636, -3.67, -8, -8, -11.027, -10.448, -8, -8, -8 \\\hline
S-2 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline

U-1 & -8, -24.368, -8, -2.226, -8, -2.649, -1.458, -4.275, -2.632, -2.409, -54.292, -2.919, -2.867, -3.325, \\
& -8, -7.464, -31.322, -2.649, -3.022, -4.018, -8, -3.313, -51.78, -50.295, -2.919, -29.011, -8, -6.055, \\
& -3.341, -8, -8, -8, -24.716, -8, -8, -8, -8, -8, -8, -8, -8  \\\hline
U-2 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, -2.415, -3.1, -6.451, \\
& -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, -8, -8, -3.794, -8, -8, -8, \\
& -8, -8, -8, -8, -8 \\\hline
  \end{tabular}
\end{table}

%\begin{table}[h]
%\centering
%\caption{Distribution of PPC codes over the optimal hidden states for successful (S) and unsuccessful (U) HMMs. Significant PPC codes are given in short form, which have log probability higher than -2.}
%\label{tab:emission_matrix}
%  \begin{tabular}{|l|l|}
%  \hline
%   \textbf{Hidden} & \textbf{Log probability of each symbol (301,LUP+, 327,OQ-TBN,101,EA,R-CT+,OQ-ECT+,AR,SUM-S,} \\ 
%   \textbf{states} & \textbf{R-CT-S,RBA-S,G-INFO+,AF,314,CQ-EF,119,R-CML+, OQ-ECML+, 307, 325, 344,CHT-,106,} \\
%    & \textbf{OQ-ECT-, HUP-O, 345, 309, R-CT-S, 114, 304, 115, 329, 330, 103, 116, 324, 323, 118, 303, 334)} \\\hline      
%S-1 & -2.896, -4.595, -4.031, -5.584, -11.383, -2.772, -1.532, -2.78, -3.816, -2.745, -8.444, -4.42, -3.485, \\
%& -2.438, -8, -5.577, -10.333, -2.216, -3.462, -3.479, -3.71, -5.615, -20.984, -7.635, -5.7, -5.694, -3.629, \\
%& -3.32, -3.195, -8, -5.007, -8, -4.636, -3.67, -8, -8, -11.027, -10.448, -8, -8, -8 \\\hline
%S-2 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
%& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
%& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline
%S-3 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
%& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
%& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline
%S-4 & -3.092, -1.344, -4.322, -4.707, -6.065, -2.205, -5.701, -2.064, -5.71, -3.778, -3.121, -6.97, -3.074, \\
%& -3.505, -8, -3.879, -3.229, -5.707, -2.708, -4.729, -5.538, -4.993, -4.116, -3.125, -8, -3.29, -4.439, \\
%& -4.414, -6.341, -8, -8, -8, -5.964, -7.009, -8, -8, -6.067, -6.071, -8, -8, -8 \\\hline
%
%U-1 & -8, -24.368, -8, -2.226, -8, -2.649, -1.458, -4.275, -2.632, -2.409, -54.292, -2.919, -2.867, -3.325, \\
%& -8, -7.464, -31.322, -2.649, -3.022, -4.018, -8, -3.313, -51.78, -50.295, -2.919, -29.011, -8, -6.055, \\
%& -3.341, -8, -8, -8, -24.716, -8, -8, -8, -8, -8, -8, -8, -8  \\\hline
%U-2 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, -2.415, -3.1, -6.451, \\
%& -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, -8, -8, -3.794, -8, -8, -8, \\
%& -8, -8, -8, -8, -8 \\\hline
%U-3 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, -2.415, -3.1, -6.451, \\
%& -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, -8, -8, -3.794, -8, -8, \\
%& -8, -8, -8, -8, -8, -8 \\\hline
%U-4 & -8, -1.021, -8, -8, -8, -3.067, -6.497, -2.993, -8, -10.079, -3.794, -8, -2.448, -8, -8, -2.415, -3.1, \\
%& -6.451, -5.018, -8, -8, -3.113, -2.407, -3.794, -8, -2.407, -8, -3.933, -7.201, -8, -8, -8, -3.794, -8, \\
%& -8, -8, -8, -8, -8, -8, -8 \\\hline
%
%  \end{tabular}
%\end{table}

\subsection*{\textit{Common patterns}}
Table~\ref{tab:common_patterns} provides examples of typical PPC sequences that frequently appear in successful and unsuccessful motivational interviews. It can be seen that most successful patterns start with a summary of the discussion or open-ended/close-ended question. After that, if adolescents show positive behavior change, the counselor immediately reflects on that positive tendency, to reinforce adolescent's desire to adhere to positive behavior change. On the other hand, general information by the counselor can lead to a negative behavior change in adolescents, even in the cases when adolescents were showing positive tendency in their previous communication. This observation can be explained by adolescents quickly loosing focus when presented with general information. \\

\begin{table}[h]
\centering
\caption{Common patterns in successful and unsuccessful motivational interviews.}
\label{tab:common_patterns}
  \begin{tabular}{|l|l|}
  \hline
   \textbf{Type} & \textbf{Pattern} \\ \hline      
successful & 328 SUM-S: Summarize $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 313 R-CML+: Reflect, \\ 
& commitment language positive $\rightarrow $ [positive commitment] \\\hline
successful & 307 SPT: Support $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 313 R-CML+: Reflect, \\
& commitment language positive $\rightarrow $ [positive commitment] \\\hline
successful & 306 CQ-EF: Closed question, Elicit Feedback $\rightarrow $ 120 HUP-O: High Uptake, other \\
& $\rightarrow $ 313 R-CML+: Reflect, commitment language positive $\rightarrow $ [positive commitment] \\\hline
unsuccessful & 311 R-CT+: Reflect, change talk positive $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 302 G-INFO+:  \\
& General Information, positive $\rightarrow $ [negative commitment] \\\hline
unsuccessful & 302 G-INFO+: General Information, positive $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $  \\
& 302 G-INFO+: General Information, positive $\rightarrow $ [negative commitment] \\\hline
unsuccessful & 305 EA: Emphasize Autonomy $\rightarrow $ 117 LUP+: Low Uptake, positive $\rightarrow $ 302 G-INFO+:  \\
& General Information, positive $\rightarrow $ [negative commitment] \\\hline
  \end{tabular}
\end{table}
